Multi-state modelling with R: the {\tt msm} package \vskip 0.2in \large{Version <<echo=FALSE,results=tex>>= cat(version) @ \vskip 0.1in <<echo=FALSE,results=tex>>= cat(format(Sys.time(), " @ }

### 2014-01-16 src/doc/msm-manual.Rnw
version 1.3

@@

\title{Multi-state modelling with R: the {\tt msm} package \vskip 0.2in
  \large{Version %
<<echo=FALSE,results=tex>>=
cat(version)
@
\vskip 0.1in
<<echo=FALSE,results=tex>>=
cat(format(Sys.time(), "%d %B, %Y"))
@
}}
\author{Christopher Jackson\\Medical Research Council Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
\date{}
%% label with date when Rnw is compiled, not when tex is compiled. should be release date of package

@@

\tableofcontents
\clearpage

@@

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
\frac{1 + (-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models. These models are illustrated in
Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, $P(t)$ is
determined from the matrix exponential, calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

@@

\begin{equation}
  \label{eq:multi:lik:exact}
  L_{i, j} =  \exp(-q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
\end{equation}
since the state is assumed to be $S(t_j)$ throughout the interval
between $t_j$ and $t_{j+1}$ with a known transition to state
$S(t_{j+1})$ at $t_{j+1}$.  \Rpackage{msm} is restricted to Markov
models, but much richer models are possible for this type of data. For
example, Putter \etal \cite{putter:mstate} discussed the
\Rpackage{mstate} software for semi-parametric multi-state models with
non-parametric baseline hazards and Cox regression.  The Markov
assumption is restrictive but necessary in general to compute a
likelihood for intermittently-observed processes.

@@

An example data set, taken from monitoring a set of heart transplant
recipients, is provided with \Rpackage{msm}.  Sharples
\etal \cite{my:cav} studied the progression of coronary allograft
vasculopathy (CAV), a post-transplant deterioration of the arterial
walls, using these data.  Risk factors and the accuracy of the
screening test were investigated using multi-state Markov and hidden
Markov models.

This data set can be made available to the current R session by
issuing the command

<<>>=
data(cav)
@

@@

The likelihood is maximised by numerical methods, which need a set of
initial values to start the search for the maximum.  For reassurance
that the true maximum likelihood estimates have been found, models
should be run repeatedly starting from different initial
values. However a sensible choice of initial values can be important
for unstable models with flat or multi-modal likelihoods.  For
example, the transition rates for a model with misclassification could
be initialised at the corresponding estimates for an approximating
model without misclassification. Initial values for a model without
misclassification could be set by supposing that transitions between
states take place only at the observation times.  If we observe
$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
package provides a function \Rfunction{crudeinits.msm} for calculating
initial values in this way.

@@

%% TODO explain about gen.inits=TRUE here

@@

By default, the data are assumed to represent snapshots of the process
at arbitrary times.  However, observations can also represent exact
times of transition, ``exact death times'', or a mixture of these. See the
\Rfunarg{obstype} argument to \Rfunction{msm}.

@@

We consider a model with just one covariate, female sex.  Out of the
622 transplant recipients, 535 are male and 87 are female.  By
default, all linear covariate effects $\beta_{rs}$ are initialised to
zero. To specify different initial values, use a \Rfunarg{covinits}
argument, as described in \Rfunction{help(msm)}.  Initial values given
in the \Rfunarg{qmatrix} represent the intensities with covariate values
set to their means in the data.

@@

@@

@@

We use an alternative quasi-Newton optimisation algorithm
\Rfunarg{(method="BFGS")} which can often be faster or more robust
than the default Nelder-Mead simplex-based algorithm.  An optional
argument \Rfunarg{initprobs} could also have been given here,
representing the vector of the probabilities of occupying each true
state at the initial observation
(equation \ref{eq:multi:hidden:matprod}).  If not given, all
individuals are assumed to be in true state 1 at their initial
observation.  If \Rfunarg{est.initprobs=TRUE} is specified, then these
probabilites are estimated as part of the model fit, using
\Rfunarg{initprobs} as initial values.  Covariate effects on these
probabilities can also be estimated using a multinomial logistic
regression model, if an \Rfunarg{initcovariates} argument is
specified.  See \Rfunction{help(msm)} for further details.

@@

\item[Odds ratios for misclassification] The function
  \Rfunction{odds.msm} gives the estimated odds ratios corresponding to
  each covariate effect on the misclassification probabilities.

<<>>=
odds.msm(cavmiscsex.msm)
@

@@

For example consider the three-state hidden Markov model described
above, with normally-distributed outcomes for states 1 and 2.
To constrain the outcome variance to be equal for states 1 and 2,
and to also constrain the effect of \Robject{x} on the outcome mean
to be equal for states 1 and 2, specify
\begin{Scode}
hconstraint = list(sd = c(1,1), x=c(1,1))
\end{Scode}

@@

Now we give an example of fitting a hidden Markov model to a real
dataset.  The data on FEV$_1$ measurements from lung transplant
recipients, described in \ref{sec:hmm:example:fev}, are provided with
the \Rpackage{msm} package in a dataset called \Robject{fev}.

<<>>=
data(fev)
@

We fit models Models 1 and 2, each with three states and common $Q$
matrix.
<<>>=
three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
@

@@

Our initial values are again 100 and 54 for the means of the
within-state distribution of \emph{underlying} FEV$_1$, and 16 and 18
for the standard errors.  This time, underlying FEV$_1$ is truncated
normal. The truncation limits \Rfunarg{lower} and \Rfunarg{upper} are
not estimated.  We take an initial measurement error standard
deviation of \Rfunarg{sderr=8}. The extra shift \Rfunarg{meanerr} in
the measurement error model is fixed to zero and not estimated.

@@

<<>>=
hmodel2 <- list(hmmMETNorm(mean=100, sd=16, sderr=8, lower=80, upper=Inf, meanerr=0),
                hmmMETNorm(mean=54, sd=18, sderr=8, lower=0, upper=80, meanerr=0),
                hmmIdent(999))

fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q, death=3, hmodel=hmodel2,
                hcovariates=list(~acute, ~acute, NULL), hcovinits = list(-8, -8, NULL),
                hconstraint = list(sderr = c(1,1), acute = c(1,1)), method="BFGS")

fev2.msm

sojourn.msm(fev2.msm)
@

@@

The following plot illustrates a trajectory of declining FEV$_1$ from
the first lung transplant recipient in this dataset.  This is produced
by the following R code.  The Viterbi algorithm is used to locate the
most likely point at which this individual moved from BOS state 1 to
BOS state 2, according to the fitted Model 2.  This is illustrated by
the vertical dotted line.  This is the point at which the individual's
lung function started to remain consistently below 80\% baseline
FEV$_1$.

<<fig=TRUE>>=
keep <- fev$ptnum==1 & fev$fev<999
plot(fev$days[keep], fev$fev[keep], type="l",
ylab=expression(paste("% baseline ", FEV[1])), xlab="Days after transplant")
vit <- viterbi.msm(fev2.msm)[keep,]
(max1 <- max(vit$time[vit$fitted==1]))
(min2 <- min(vit$time[vit$fitted==2]))
abline(v = mean(max1,min2), lty=2)
text(max1 - 500, 50, "STATE 1")
text(min2 + 500, 50, "STATE 2")
@

@@

<<>>=
oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
                   c(0, 0, 0.202, 0.081),
                   c(0, 0, 0, 0.126),
                   c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
cavmisc.msm
@

@@

Your new distribution will be available to use in the \Rfunarg{hmodel}
argument to \Rfunction{msm}, as, for example
\begin{Scode}
hmodel = list(..., hmmNewDist(location = 0, scale = 1), ...)
\end{Scode}
If your distribution may be of interest to others, ask me nicely
(\texttt{chris.jackson@mrc-bsu.cam.ac.uk}) to include it in a future
release.

@@

>>>>

%% version <- 1.3

@@

\title{Multi-state modelling with R: the {\tt msm} package \vskip 0.2in
  \large{Version %
<<echo=FALSE,results=tex>>=
cat(version)
@
\vskip 0.1in
<<echo=FALSE,results=tex>>=
cat(format(Sys.time(), "%d %B, %Y"))
@
}}
\author{Christopher Jackson\\MRC Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
\date{}
%% label with date when Rnw is compiled, not when tex is compiled. should be release date of package

@@

@@

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
1 + \frac{(-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models, illustrated in
Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
structure on any number of states in principle, $P(t)$ is
determined from the matrix exponential.  This is calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

@@

\begin{equation}
  \label{eq:multi:lik:exact}
  L_{i, j} =  \exp(q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
\end{equation}
since the state is assumed to be $S(t_j)$ throughout the interval
between $t_j$ and $t_{j+1}$ with a known transition to state
$S(t_{j+1})$ at $t_{j+1}$.  \Rpackage{msm} is restricted to Markov
models, but much richer models are possible for this type of data. For
example, Putter \etal \cite{putter:mstate} discussed the
\Rpackage{mstate} software for semi-parametric multi-state models with
non-parametric baseline hazards and Cox regression.  The Markov
assumption is restrictive but necessary in general to compute a
likelihood for intermittently-observed processes.

@@

An example data set, taken from monitoring a set of heart transplant
recipients, is provided with \Rpackage{msm}.  (Note: since \Rpackage{msm}
version 1.3, the command \Rfunction{data(cav)} is no longer needed to
load the data --- it is now ``lazy-loaded'' when required). Sharples
\etal \cite{my:cav} studied the progression of coronary allograft
vasculopathy (CAV), a post-transplant deterioration of the arterial
walls, using these data.  Risk factors and the accuracy of the
screening test were investigated using multi-state Markov and hidden
Markov models.

@@

The likelihood is maximised by numerical methods, which need a set of
initial values to start the search for the maximum.  For reassurance
that the true maximum likelihood estimates have been found, models
should be run repeatedly starting from different initial
values. However a sensible choice of initial values can be important
for unstable models with flat or multi-modal likelihoods.  For
example, the transition rates for a model with misclassification could
be initialised at the corresponding estimates for an approximating
model without misclassification. Initial values for a model without
misclassification could be set by supposing that transitions between
states take place only at the observation times.  If we observe
$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$.

Such default initial values can be used by supplying
\Rfunarg{gen.inits=TRUE} in the call to \Rfunction{msm} below, along
with a \Rfunarg{qmatrix} whose non-zero entries represent the allowed
transitions of the model.  Alternatively the function \Rfunction{crudeinits.msm} 
could be used to get the  amatrix of initial values as follows.  
These methods are only available for non-hidden Markov models.

@@

@@

By default, the data are assumed to represent snapshots of the process
at arbitrary times.  However, observations can also represent exact
times of transition, ``exact death times'', or a mixture of these. See the
\Rfunarg{obstype} argument to \Rfunction{msm}.  Note you do not always have to specify \Rfunarg{death},
even if there are states which correspond to death.  This is
only used for states whose entry time is observed in this specific way.

@@

We consider a model with just one covariate, female sex.  Out of the
622 transplant recipients, 535 are male and 87 are female.  By
default, all linear covariate effects $\beta_{rs}$ are initialised to
zero. To specify different initial values, use a \Rfunarg{covinits}
argument, as described in \Rfunction{help(msm)}.  Initial values given
in the \Rfunarg{qmatrix} represent the intensities with covariate values
set to their means in the data.  In the following model, all transition
intensities are modelled in terms of sex.

@@

Different transition rates may be easily modelled on different
covariates by specifying a named list of formulae as the
\Rfunarg{covariates} argument.  Each element of the list has a name
identifying the transition.  In the model below, the transition rate
from state 1 to state 2 and the rate from state 3 to state 4 are each
modelled on sex as a covariate, but no other intensities have
covariates on them.

\paragraph{Model 2a: transition-specific covariates}
<<eval=FALSE>>=
cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                  qmatrix = twoway4.q, death = 4,
                  covariates = list("1-2" = ~ sex, "3-4" = ~sex) )
@

@@

\item[\Rfunction{optim} "function cannot be evaluated at initial parameters"]
  To diagnose this problem, run \Rfunction{msm} again with
  \Rfunarg{fixedpars=TRUE} set, to calculate the -2 log-likelihood at
  the initial values.  This will probably be \Robject{Inf}. To show
  the contributions of individual subjects to the overall log
  likelihood, call \Rfunction{logLik.msm(x, by.subject=TRUE)}, where
  \Robject{x} is the fitted model object.  If only a few subjects give
  an infinite log-likelihood, then you can check whether their state
  histories are particularly unusual and conflict with the model.  For
  example, they might appear to make unusually large jumps between
  states in short periods of time.  For models with misclassification,
  note that the default true initial state distribution
  \Rfunarg{initprobs} puts all individuals in true state 1 at their
  first observation.  If someone starts in a much higher state, this
  may result in an infinite log-likelihood, and changing
  \Rfunarg{initprobs} would be sensible.

@@

We use an alternative quasi-Newton optimisation algorithm
\Rfunarg{(method="BFGS")} which can often be faster or more robust
than the default Nelder-Mead simplex-based algorithm.  An optional
argument \Rfunarg{initprobs} could also have been given here,
representing the vector of the probabilities of occupying each true
state at the initial observation
(equation \ref{eq:multi:hidden:matprod}).  This can also be a matrix
with number of rows equal to the number of subjects, if these
probabilities are subject-dependent and known. If not given, all
individuals are assumed to be in true state 1 at their initial
observation.  If \Rfunarg{est.initprobs=TRUE} is specified, then these
probabilites are estimated as part of the model fit, using
a vector \Rfunarg{initprobs} as initial values.  Covariate effects on these
probabilities can also be estimated using a multinomial logistic
regression model, if an \Rfunarg{initcovariates} argument is
specified.  See \Rfunction{help(msm)} for further details.

@@

\item[Odds ratios for misclassification] The function
  \Rfunction{odds.msm} would give the estimated odds ratios corresponding to
  each covariate effect on the misclassification probabilities.

\begin{Scode}
odds.msm(cavmiscsex.msm)
\end{Scode}

@@

For example, consider the three-state hidden Markov model described
above, with normally-distributed outcomes for states 1 and 2.
To constrain the outcome variance to be equal for states 1 and 2,
and to also constrain the effect of \Robject{x} on the outcome mean
to be equal for states 1 and 2, specify
\begin{Scode}
hconstraint = list(sd = c(1,1), x=c(1,1))
\end{Scode}

Parameters of the
outcome distributions may also be constrained within specific ranges.
If chosen carefully, this may improve identifiability of hidden Markov states.
For example to constrain the mean for state 1 to be between 80 and 110,
and the mean for state 2 to be between 50 and 80, specify
\begin{Scode}
hranges = list(mean=list(lower=c(80,50), upper=c(110,80)))
\end{Scode}
Maximum likelihood estimation is then performed on the appropriate log
or logit-transformed scale so that these constraints are
satisfied. See the \Rfunction{msm} help page for further details.
Note that initial values should be strictly within the ranges, and not
on the range boundary.

@@

Now we give an example of fitting a hidden Markov model to a real
dataset.  The data on FEV$_1$ measurements from lung transplant
recipients, described in \ref{sec:hmm:example:fev}, are provided with
the \Rpackage{msm} package in a dataset called \Robject{fev}.
We fit models Models 1 and 2, each with three states and common $Q$
matrix.
<<>>=
three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
@

@@

Our initial values are 90 and 54 for the means of the
within-state distribution of \emph{underlying} FEV$_1$, and 16 and 18
for the standard errors.  This time, underlying FEV$_1$ is truncated
normal. The truncation limits \Rfunarg{lower} and \Rfunarg{upper} are
not estimated.  We take an initial measurement error standard
deviation of \Rfunarg{sderr=8}. The extra shift \Rfunarg{meanerr} in
the measurement error model is fixed to zero and not estimated.

@@

The convergence of maximum likelihood estimation in this example is
particularly sensitive to the optimisation method and options, initial values,
the unit of the time variable and whether covariates are centered,
probably because the likelihood surface is irregular near to the
true maximum.

\begin{Scode}
hmodel2 <- list(hmmMETNorm(mean=90, sd=16, sderr=8,
                           lower=80, upper=Inf, meanerr=0),
                hmmMETNorm(mean=54, sd=18, sderr=8,
                           lower=0, upper=80, meanerr=0),
                hmmIdent(999))

fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
                death=3, hmodel=hmodel2,
                hcovariates=list(~acute, ~acute, NULL),
                hcovinits = list(-8, -8, NULL),
                hconstraint = list(sderr = c(1,1), acute = c(1,1)),
                control=list(maxit=10000), center=TRUE)
\end{Scode}

@@

The following code will create a plot that illustrates a trajectory of declining FEV$_1$ from
the first lung transplant recipient in this dataset.  The Viterbi algorithm is used to locate the
most likely point at which this individual moved from BOS state 1 to
BOS state 2, according to the fitted Model 2.  This is illustrated by
a vertical dotted line.  This is the point at which the individual's
lung function started to remain consistently below 80\% baseline
FEV$_1$.

\begin{Scode}
keep <- fev$ptnum==1 & fev$fev<999
plot(fev$days[keep], fev$fev[keep], type="l",
ylab=expression(paste("% baseline ", FEV[1])), xlab="Days after transplant")
vit <- viterbi.msm(fev2.msm)[keep,]
(max1 <- max(vit$time[vit$fitted==1]))
(min2 <- min(vit$time[vit$fitted==2]))
abline(v = mean(max1,min2), lty=2)
text(max1 - 500, 50, "STATE 1")
text(min2 + 500, 50, "STATE 2")
\end{Scode}

\includegraphics{figures/fev_viterbi}

@@

\begin{Scode}
oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
                   c(0, 0, 0.202, 0.081),
                   c(0, 0, 0, 0.126),
                   c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), 
                                    hmmCat(c(0.1, 0.8, 0.1, 0)), 
                                    hmmCat(c(0, 0.1, 0.9, 0)), 
                                    hmmIdent(4)),
                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, 
                     method="BFGS")
cavmisc.msm
\end{Scode}

@@

Your new distribution will be available to use in the \Rfunarg{hmodel}
argument to \Rfunction{msm}, as, for example
\begin{Scode}
hmodel = list(..., hmmNewDist(location = 0, scale = 1), ...)
\end{Scode}
If your distribution may be of interest to others, ask me
(\texttt{chris.jackson@mrc-bsu.cam.ac.uk}) to include it in a future
release.

@@

Development versions of \Rpackage{msm} are to be found on R-Forge,
\url{http://msm.r-forge.r-project.org/}.  These are often more recent
than the version released on CRAN, so if you think you have found a
bug, then please check first to see whether it has been fixed on the
R-Forge version.
